Combined annotation exploration for SCPCP000006

### 2024-11-15 analyses/cell-type-wilms-tumor-06/notebook/07_combined_annotation_across_samples_exploration.Rmd
fix 07 typos and table formatting, and regenerate notebook

first   level annotation | second level   annotation | selection of   the cells | marker genes   for validation | cnv validation
-- | -- | -- | -- | --
normal | endothelial | compartment == "endothelium"     & predicted.score > pred.thr     & cnv_score < cnv.thr | WVF | no   cnv
normal | immune | compartment == "immune"     & predicted.score > pred.thr     & cnv_score < cnv.thr | PTPRC, CD163, CD68 | no cnv
normal | kidney | cell_type %in% c("kidney cell",   "kidney epithelial", "podocyte")     & predicted.score > pred.thr     & cnv_score < cnv.thr | CDH1, PODXL, LTL | no cnv
normal | stroma | compartment == "stroma"     & predicted.score > pred.thr     & cnv_score < cnv.thr | VIM | no cnv
cancer | stroma | compartment == "stroma"     & cnv_score > cnv.thr | VIM | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
cancer | blastema | compartment ==   "fetal_nephron"     & cell_type == "mesenchymal cell"     & cnv_score > cnv.thr | CITED1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
cancer | epithelial | compartment ==   "fetal_nephron"     & cell_type != "mesenchymal cell"     & cnv_score > cnv.thr | CDH1 | proportion_cnv_chr     -1     -4     -11     -16     -17     -18
unknown | - | the rest of the cells | - | proportion_cnv_chr     -1     -4     -11     -16     -17     -18

@@

### Global cnv score

@@

- _normal_ cells can be observe in all four compartments (`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have cnv.
We only allow a bit of flexibility in terms of cnv profile for immune and endothelium cells that have a high predicted score.
Indeed, we know that false positive cnv can be observed in a cell type specific manner.

The threshold used for the `predicted.score` is defined as a parameter of this notebook as `r params$predicted.celltype.threshold`.
The threshold used for the identification of cnv is also defined in the params of the notebook as `r params$cnv_threshold`.

- _cancer_ cells are either from the `stroma` or `fetal nephron` compartments and must have at least few cnv.

@@

- _blastema_,: we defined as _bastema_ every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`.
We know that these _mesenchymal_ cells are cells from the cap mesenchyme that are not expected to be in a mature kidney.
These blastema cells should express higher _CITED1_.

@@

### Validation cancer versus normal based on the cnv profile

@@

#### Immune, _PTPRC_ expression

@@

#### Endothelium, _VWF_ expression

@@

#### Stroma, _Vimentin_ expression

@@

#### Stroma, _COL6A3_ expression

@@

#### Stroma, _THY1_ expression

@@

#### Blastema, _CITED1_ expression

@@

#### Blastema, _NCAM1_ expression

@@

#### stemness marker (blastema and primitive epithelium), _SIX2_ expression

@@

#### Epithelium, _CDH1_ expression

@@

#### Epithelium, _PODXL_ expression

@@

- The heatmaps of cnv proportion and marker genes support our annotations, but signals with some marker genes are very low.
Also, there is no universal marker for each entity of Wilms tumor that cover all tumor cells from all patient.
This makes the validation of the annotations quite difficult.

@@

- In each histology (i.e. epithelial and stroma), the distinction between cancer and non cancer cell is difficult (as expected).
In this analysis, we suggested to rely on the cnv score to assess the normality of the cell.
Here again, we could try to run differential expression analysis and compare epithelial (resp. stroma) cancer versus non-cancer cells across patient, aiming to find a share transcriptional program allowing the classification cancer versus normal.

- In our annotation, we haven't taken into account the favorable/anaplastic status of the sample.
However, as anaplasia can occur in every (but do not has to) wilms tumor histology, I am not sure how to integrate the information into the annotation.

- This notebook could be finally rendered using different parameters, i.e. threshold for the cnv score and predicted score to use.

>>>>

| first level annotation | second level annotation | selection of the cells | marker genes for validation | CNV validation |
| ---------------------- | ----------------------- | ---------------------- | --------------------------- | --------------- |
| normal                 | endothelial             | `compartment == "endothelium" & predicted.score > pred.thr  & cnv_score < cnv.thr` | `VWF`| no CNV |
| normal                 | immune                  | `compartment == "immune" & predicted.score > pred.thr & cnv_score < cnv.thr` | `PTPRC`, `CD163`, `CD68`| no CNV |
| normal                 | kidney                  | `cell_type %in% c("kidney cell","kidney epithelial", "podocyte") & predicted.score > pred.thr & cnv_score < cnv.thr` | `CDH1`, `PODXL`, `LTL`| no CNV |
| normal                 | stroma                  | `compartment == "stroma" & predicted.score > pred.thr & cnv_score < cnv.thr`| `VIM`| no CNV |
| cancer                 | stroma                  | `compartment == "stroma" & cnv_score > cnv.thr` | `VIM`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | blastema                | `compartment == "fetal_nephron" & cell_type == "mesenchymal cell" & cnv_score > cnv.thr`  | `CITED1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| cancer                 | epithelial              | `compartment == "fetal_nephron" & cell_type != "mesenchymal cell" & cnv_score > cnv.thr`  | `CDH1`| `proportion_cnv_chr: 1, 4, 11, 16, 17, 18` |
| unknown                | -                       | the rest of the cells  | -  | -|

@@

### Global CNV score

@@

- _normal_ cells can be observe in all four compartments (`endothelium`, `immune`, `stroma` or `fetal nephron`) and do not have CNV
We only allow a bit of flexibility in terms of CNV profile for immune and endothelium cells that have a high predicted score.
Indeed, we know that false positive CNV can be observed in a cell type specific manner.

The threshold used for the `predicted.score` is defined as a parameter of this notebook as `r params$predicted.celltype.threshold`.
The threshold used for the identification of CNV is also defined as the notebook parameter `r params$cnv_threshold`.

- _cancer_ cells are either from the `stroma` or `fetal nephron` compartments and must have at least few CNV

@@

- _blastema_,: we defined as _blastema_ every cancer cell that has a `fetal_kidney_predicted.cell_type == mesenchymal cell`.
We know that these _mesenchymal_ cells are cells from the cap mesenchyme that are not expected to be in a mature kidney.
These blastema cells should express higher _CITED1_.

@@

### Validation cancer versus normal based on the CNV profile

@@

#### Immune, `PTPRC` expression

@@

#### Endothelium, `VWF` expression

@@

#### Stroma, `Vimentin` expression

@@

#### Stroma, `COL6A3` expression

@@

#### Stroma, `THY1` expression

@@

#### Blastema, `CITED1` expression

@@

#### Blastema, `NCAM1` expression

@@

#### stemness marker (blastema and primitive epithelium), `SIX2` expression

@@

#### Epithelium, `CDH1` expression

@@

#### Epithelium, `PODXL` expression

@@

- The heatmaps of CNV proportion and marker genes support our annotations, but signals with some marker genes are very low.
Also, there is no universal marker for each entity of Wilms tumor that cover all tumor cells from all patient.
This makes the validation of the annotations quite difficult.

@@

- In each histology (i.e. epithelial and stroma), the distinction between cancer and non cancer cell is difficult (as expected).
In this analysis, we suggested to rely on the CNV score to assess the normality of the cell.
Here again, we could try to run differential expression analysis and compare epithelial (resp. stroma) cancer versus non-cancer cells across patient, aiming to find a share transcriptional program allowing the classification cancer versus normal.

- In our annotation, we haven't taken into account the favorable/anaplastic status of the sample.
However, as anaplasia can occur in every (but do not has to) Wilms tumor histology, I am not sure how to integrate the information into the annotation.

- This notebook could be finally rendered using different parameters, i.e. threshold for the CNV score and predicted score to use.
